---
title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
header-includes:
    \usepackage{placeins}
output:  
  bookdown::html_document2:
    css: "style.css" #style for the HTML document
  bookdown::word_document2:
    fig_caption: yes #figure caption
    keep_md: yes
  bookdown::pdf_document2:
    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
    pandoc_args: --listings #calls the listings package to fit code in the page margins
    keep_tex: yes #keep LaTeX file for submission
    fig_caption: yes #allows captions in figures
    extra_dependencies:
      subfig: null #allows for subfigures
      breqn: null #line breaks for long equations
      caption: ["font={small}"] #size of the figure captions
      float: null #allows for control of placement of figures
    includes:
      in_header: preamble.sty #additional LaTeX formatting
csl: elsevier-with-titles.csl #style for references
bibliography: refs.bib #references
link-citations: yes #adds links to the citations
'': default
---


# Abstract

In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, a _linear mixed model_ (LMEM). Although LMEMs are less restrictive that rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear. In contrast, generalized additive models (GAMs), are a class of models that relax the linearity assumption and allow the data to determine the fit of the model while permitting missing observations and different correlation structures, thereby being an excellent choice to analyze non-linear longitudinal data. This paper summarizes the limitations of rm-ANOVA and LEMEs, presents the basic theory of GAMs, and demonstrates their implementation in $\textsf{R}$ via the package _mgcv_ using simulated data that follows longitudinal trends reported in biomedical literature. To promote reproducibility in biomedical research, the code and data used to generate this paper are available at:________.


```{r libraries, message = FALSE,include=FALSE}
#the packages needed to run the code

library(patchwork)
library(tidyverse)
library(mvnfast)
library(nlme)
library(mgcv)
library(gratia)
library(viridis)
set.seed(1) #set seed for reproducibility

knitr::opts_chunk$set(fig.pos = 'H', out.extra = "",error=TRUE)

#global themes for plotting
thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting

#thm<-scale_color_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting
#thm1<-scale_fill_ghibli_d("MarnieMedium2",direction=-1) #global theme for plotting

#this function will remove the default "author information" that is the Pandoc LaTeX template. Otherwise, and additional "and" will appear at the end of the author list.Or, the custom template that is in the YAML header and that has the "author" line deleted may be used as well. An explanation in detail is provided at:
#https://github.com/rstudio/bookdown/issues/1158
remove_author <- function(x) {
  # identify empty author line
  i <- grep("^\\\\author\\{\\}$", x)
  # be sure it is the one pandoc inserts
  if(length(i) != 0 && grepl('^\\\\date\\{', x[i+1])) x <- x[-i]
  x
}
options(bookdown.post.latex = remove_author)
```


```{r child='01-Background.Rmd'}
```

\FloatBarrier

```{r child='02-Challenges.Rmd'}
```

\FloatBarrier

```{r child='03-GAM Theory.Rmd'}
```

\FloatBarrier

```{r child='04-Longitudinal analysis with GAMs.Rmd'}
```

\FloatBarrier

```{r child='05-Conclusion.Rmd'}
```

\FloatBarrier

***


\newpage
# References

<div id="refs"></div>


```{r child='06-Appendix.Rmd'}
```






