
# The analyisis of longitudinal biomedical data using GAMs{#longitudinal-GAMs}

The previous sections provided the basic framework to understand the GAM framework and how these models are more advantageous to analyze non-linear longitudinal data when compared to rm-ANOVA or LMEMs. This section will use simulation to present the practical implementation of GAMs for longitudinal biomedical data using $\textsf{R}$ and the package `mgcv`. The code for the simulated data and figures, and a brief guide for model selection and diagnostics appear in the Appendix.

## Simulated data

The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors (Figure 3C in Vishwanath et. al.[@vishwanath2009]). In the paper, diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10). In the "Treatment" group (chemotherapy) an increase in $\mbox{StO}_2$ is observed through time, while a decrease is seen in the "Control" (saline) group. Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (matching the SD in the original paper). 
The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot)A.

## An interaction GAM for longitudinal data

An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination. In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the _time_ effect to account for the repeated measures over time.  Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in Hefley [@hefley2017]. 

For the data in Figure \@ref(fig:sim-smooth-plot)A the main effect of interest is how $\mbox{StO}_2$ changes over time for each treatment. To estimate this, the model incorporates separate smooths for _Group_ and _Day_, respectively. The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (*Day*) and the other is a factor (*Group*). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the group means. Using $\textsf{R}$ and the package `mgcv` the model syntax is:

```{r, eval = FALSE}
gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
```


This syntax specifies that `gam_02` (named this way so it matches the model workflow from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (`StO2_sim`) is modeled using independent smooths over `Day` for each `Group` (the parenthesis preceded by `s`) using four basis functions (plus the intercept). The smooth is constructed by default using TPRS, but other splines can be used if desired, including Gaussian process smooths [@simpson2018] (a description of all the available smooths can be found by typing `?mgcv::smooth.terms` in the Console). 

The parametric term `Group` is added to quantify overall mean differences in the effect of treatment between groups. Although the default `method` used to estimate the smoothing parameters in _mgcv_ is generalized cross validation (GCV), Wood[@wood2017] showed the restricted maximum likelihood (REML) to be more resistant to overfitting while also easing the quantification of uncertainty in the smooth parameters; therefore in this manuscript REML is always used for smooth parameter estimation. An additional argument (`family`) allows to specify the expected distribution of the response, but it is not used in this model because we expect a normally-distributed response (which is the default `family` in _mgcv_).

When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of $\mbox{StO}_2$ for each group across time (Figure \@ref(fig:sim-smooth-plot)B). Model diagnostics can be obtained using the `gam.check` function, and the function `appraise` from the package _gratia_ [@gratia]. A guide for model selection and diagnostics is in the [Appendix](#workflow), and an in-depth analysis can be found in Wood [@wood2017] and Harezlak [@harezlak2018]. 

One question that might arise at this point is "what is the fit that an rm-ANOVA model produces for the simulated data?". The rm-ANOVA model, which corresponds to Equation \@ref(eq:linear-model) is presented in Figure \@ref(fig:sim-smooth-plot)C. This is a typical case of model misspecification: The slopes of each group are different, which would lead to a _p-value_ indicating significance for the treatment and time effects, but the model is not capturing the changes that occur at days 2 and between days 5 and 7, whereas the GAM model is able to reliably estimate the trend over all timepoints (Figure \@ref(fig:sim-smooth-plot)B) .

```{r,sim-tumor-response,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
dat<-tibble(StO2=c(4,27,3,2,0.5,7,4,50,45,56),
            Day=rep(c(0,2,5,7,10),times=2),
            Group=as.factor(rep(c("Control","Treatment"),each=5))
            )
#alpha for ribbon
al<-0.8


#This function simulates data for the tumor data using default parameters of 10 observations per time point,and Standard deviation (sd) of 5%.
#Because physiologically StO2 cannot go below 0%, data is  generated with a cutoff value of 0.0001 (the "StO2_sim")

simulate_data <- function(dat, n = 10, sd = 5) {
    dat_sim <- dat %>%
        slice(rep(1:n(), each = n)) %>%
        group_by(Group, Day) %>%
        mutate(
               StO2_sim = pmax(rnorm(n, StO2, sd), 0.0001),
               subject=rep(1:10),
               subject=factor(paste(subject, Group, sep = "-"))
               ) %>%
        ungroup()

    return(dat_sim)
}


#subject = factor(paste(subject, treatment, sep = "-")))

n <- 10 #number of observations
sd <- 10 #approximate sd from paper
df <- 6
dat_sim <- simulate_data(dat, n, sd)

#plotting simulated data
f2<-ggplot(dat_sim, aes(x = Day, 
                        y = StO2_sim, 
                        color = Group,
                        group=subject)) +
    geom_point(show.legend=FALSE,
               size=1.5,
               alpha=0.6)+
    geom_line(size=0.6, alpha=0.4,show.legend=FALSE)+
    geom_line(aes(x = Day, 
               y = StO2,
               color=Group),
           size=1.5,
           data=dat,
           inherit.aes=FALSE,
              show.legend = FALSE)+
  labs(y=expression(atop(StO[2],'(simulated)')))+
  theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    thm1
  

```



```{r,smooth-plot,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}

#GAM for StO2

gam_02 <- gam(StO2_sim ~ Group+s(Day, by = Group, k = 5),
            method='REML',
            data  = dat_sim)



#linear model
lm1<-lm(StO2_sim ~ Day + Group + Day * Group, data = dat_sim)

## point-wise interval
ci <- confint(gam_02, parm = "s(Day)", partial_match = TRUE, type = "confidence")
## simultaneous interval
si <- confint(gam_02, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)


# mean shift for Treatment group
const <- coef(gam_02)[2]

#pointwise confidence interval
ci <- ci %>%
mutate(est = case_when(Group == "Treatment" ~ est + const,
TRUE ~ est),
lower = case_when(Group == "Treatment" ~ lower + const,
TRUE ~ lower),
upper = case_when(Group == "Treatment" ~ upper + const,
TRUE ~ upper))

#simultaneous interval
si <- si %>%
mutate(est = case_when(Group == "Treatment" ~ est + const,
TRUE ~ est),
lower = case_when(Group == "Treatment" ~ lower + const,
TRUE ~ lower),
upper = case_when(Group == "Treatment" ~ upper + const,
TRUE ~ upper))


#creates a dataframe using the length of the covariates for rm-ANOVA
lm_predict<-expand_grid(Group = factor(c("Control", "Treatment")),
                         Day = c(0:10),
                        subject=factor(rep(1:10)),
                          )
lm_predict$subject<-factor(paste(lm_predict$subject, lm_predict$Group, sep = "-"))

#using lm
lm_predict<-lm_predict%>%
    mutate(fit = predict(lm1,lm_predict,se.fit = TRUE,type='response')$fit,
           se.fit = predict(lm1, lm_predict,se.fit = TRUE,type='response')$se.fit)

#plot smooths and pointwise and simulatenous confidence intervals for GAM
f3<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
geom_line(lwd = 1) +
geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
inherit.aes = FALSE, alpha = 0.7,
show.legend=FALSE) +
geom_ribbon(data = si,
mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
inherit.aes = FALSE, alpha = 0.3,
show.legend=FALSE)+
    geom_point(data=dat_sim, aes(x = Day, 
                        y = StO2_sim, 
                        color = Group), 
                        size=1.5,
                        alpha=0.6, 
               inherit.aes = FALSE,
               show.legend = FALSE)+
    geom_line(data=si,aes(Day,upper,color=Group), size=0.8, alpha=0.7)+
    geom_line(data=si,aes(Day,lower,color=Group), size=0.8, alpha=0.7)+
    labs(y=expression(atop(StO[2],'(complete observations)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+thm1

 
#plot linear fit for rm-ANOVA
f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),fill=Group),
              alpha=al,
              data=lm_predict,
              show.legend = FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=lm_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(paste('StO'[2],' (simulated)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+thm1

```


Because GAMs do not require equally-spaced or complete observations for all subjects (as rm-ANOVA does), they are advantageous to analyze longitudinal data where missingness exists. The rationale behind this is that GAMs are able to pick the trend in the data even when some observations are missing. However, this usually causes the resulting smooths to have wider confidence intervals and less ability to pick certain trends. Consider the  simulated $\mbox{StO}_2$ values from Figure \@ref(fig:sim-smooth-plot)B. If 40% of the observations are randomly deleted and the same interaction GAM (`gam_02`) is fitted, the resulting smooths appear in (Figure \@ref(fig:sim-smooth-plot)D). Note that the model is still able to show a different trend for each group, but with a somewhat more linear profile in some areas. 

Additionally, note that in Figures \@ref(fig:sim-smooth-plot)B,D we show two CIs for each of the fitted smooths (shaded regions). The across the function CIs are represented by the narrow regions and because the simultaneous CIs contain the whole function on a nominal value, we expect them to be wider than the across the function CI, resulting in the wide shaded regions. For the dataset with missing observations, the smooths CIs overlap during the first 3 days because the estimates are less robust with fewer data points, and the trend is less pronounced than in the full dataset. However, the overall trend of the data is picked by the model in both cases, with as few as 4 observations per group at certain time points.

```{r,missing-data,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
#missing data
#create a sequence of 40 random numbers between 1 and 100, these numbers will
#correspond to the row numbers to be randomly erased from the original dataset
missing <- sample(1:100, 40)

#create a new dataframe from the simulated data with 40 rows randomly removed, keep the missing values as NA

ind <- which(dat_sim$StO2_sim %in% sample(dat_sim$StO2_sim, 40))

#create a new dataframe, remove the StO2 column
dat_missing <- dat_sim[,-1]

#add NAs at the ind positions
dat_missing$StO2_sim[ind]<-NA 

#Count the number of remaining observations per day (original dataset had 10 per group per day)
dat_missing %>%
    group_by(Day,Group) %>%
    filter(!is.na(StO2_sim))%>%
  count(Day)


#the same model used for the full dataset
mod_m1 <- gam(StO2_sim ~ Group+s(Day,by=Group,k=5), data  = dat_missing,family=scat)
#appraise the model
appraise(mod_m1)


ci <- confint(mod_m1, parm = "s(Day)", partial_match = TRUE, type = "confidence")
## simultaneous interval
si <- confint(mod_m1, parm = "s(Day)", type = "simultaneous", partial_match = TRUE)


# mean shift for group 2
const <- coef(mod_m1)[2]

ci <- ci %>%
mutate(est = case_when(Group == "Treatment" ~ est + const,
TRUE ~ est),
lower = case_when(Group == "Treatment" ~ lower + const,
TRUE ~ lower),
upper = case_when(Group == "Treatment" ~ upper + const,
TRUE ~ upper))

si <- si %>%
mutate(est = case_when(Group == "Treatment" ~ est + const,
TRUE ~ est),
lower = case_when(Group == "Treatment" ~ lower + const,
TRUE ~ lower),
upper = case_when(Group == "Treatment" ~ upper + const,
TRUE ~ upper))


f6<-ggplot(ci, aes(x = Day, y = est, group = smooth)) +
geom_line(lwd = 1) +
geom_ribbon(data = ci, mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill = Group),
inherit.aes = FALSE, alpha = 0.7,
show.legend=FALSE) +
geom_ribbon(data = si,
mapping = aes(ymin = lower, ymax = upper, x = Day, group = smooth,fill =Group),
inherit.aes = FALSE, alpha = 0.4,
show.legend=TRUE)+
    geom_line(data=si,aes(Day,upper,color=Group), size=0.8, alpha=0.7)+
    geom_line(data=si,aes(Day,lower,color=Group), size=0.8, alpha=0.7)+
    geom_point(data=dat_missing, aes(x = Day, 
                        y = StO2_sim, 
                        color = Group), 
                            size=1.5,
                        alpha=0.6, 
               inherit.aes = FALSE)+
    labs(y=expression(atop(StO[2],'(missing observations)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22))+
        thm1

```

(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors. **A**: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment. Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point. **B**: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment. Lines represent trends for each group, shaded regions are 95% across the function (narrow region) and simultaneous (wide region) confidence intervals. **C**: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time. **D**: Smooths for the GAM model for the simulated data with missing observations (40%). Lines represent trends for each group, shaded regions are 95%  across the function (narrow region) and simultaneous (wide region) confidence intervals.

```{r,sim-smooth-plot,fig.width=12, fig.height=7, out.width='100%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}


mult_plot<-f2+f3+f4+f6+
   plot_annotation(tag_levels='A')&
   ylim(c(-20,75)) &
  theme(
     text=element_text(size=18)
     )

mult_plot

```



```{r,pairwise-comp,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}

##Pairwise comparisons


difference_pointwise <- function(f1, f2, smooth, by_var, smooth_var, data, Xp, V, coefs, nrep = 1000) {
  ## make sure f1 and f2 are characters
  f1 <-  as.character(f1)
  f2 <-  as.character(f2)
  cnames <- colnames(Xp)
  ## columns of Xp associated with pair of smooths
  c1 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f1), cnames, fixed = TRUE)
  c2 <- grepl(gratia:::mgcv_by_smooth_labels(smooth, by_var, f2), cnames, fixed = TRUE)
  ## rows of Xp associated with pair of smooths
  r1 <- data[[by_var]] == f1
  r2 <- data[[by_var]] == f2

  ## difference rows of Xp for pair of smooths
  X <- Xp[r1, ] - Xp[r2, ]

  ## zero the cols related to other splines
 # X[, ! (c1 | c2)] <- 0

  ## zero out the parametric cols
  #X[, !grepl('^s\\(', cnames)] <- 0

  ## compute difference
  sm_diff <- drop(X %*% coefs)
  se <- sqrt(rowSums((X %*% V) * X))
  nr <- NROW(X)

  ## Calculate posterior simulation for smooths
  coefs_sim <- t(rmvn(nrep, rep(0, nrow(V)), V))
  rownames(coefs_sim) <- rownames(V)
  simDev <- X %*% coefs_sim
  absDev <- abs(sweep(simDev, 1, se, FUN = "/"))
  masd <- apply(absDev, 2, max)
  crit_s <- quantile(masd, prob = 0.95, type = 8)


  out <- list(smooth = rep(smooth, nr), by = rep(by_var, nr),
              level_1 = rep(f1, nr),
              level_2 = rep(f2, nr),
              diff = sm_diff, se = se,
              lower_s = sm_diff - crit_s * se,
              upper_s = sm_diff + crit_s*se)

  out <- new_tibble(out, nrow = NROW(X), class = "difference_smooth")
  ## Only need rows associated with one of the levels
  out <- bind_cols(out, data[r1, smooth_var])

  out
}

#does both ci and si
difference_smooths <- function(model,
                                 smooth,
                                 n = 100,
                                 ci_level = 0.95,
                                 newdata = NULL,
                                 partial_match = TRUE,
                                 unconditional = FALSE,
                                 frequentist = FALSE,
                                 nrep = 10000,
                                 include_means = TRUE,
                                 ...) {
  if (missing(smooth)) {
    stop("Must specify a smooth to difference via 'smooth'.")
  }

  # smooths in model
  S <- gratia::smooths(model) # vector of smooth labels - "s(x)"
  # select smooths
  select <-
    gratia:::check_user_select_smooths(smooths = S, select = smooth,
                                       partial_match = partial_match)#,
  # model_name = expr_label(substitute(object)))
  sm_ids <- which(select)
  smooths <- gratia::get_smooths_by_id(model, sm_ids)
  sm_data <- map(sm_ids, gratia:::smooth_data,
                 model = model, n = n, include_all = TRUE)
  sm_data <- bind_rows(sm_data)
  by_var <- by_variable(smooths[[1L]])
  smooth_var <- gratia:::smooth_variable(smooths[[1L]])
  pairs <- as_tibble(as.data.frame(t(combn(levels(sm_data[[by_var]]), 2)),
                                   stringsAsFactor = FALSE))
  names(pairs) <- paste0("f", 1:2)

  Xp <- predict(model, newdata = sm_data, type = "lpmatrix")
  V <- gratia:::get_vcov(model, unconditional = unconditional,
                         frequentist = frequentist)
  coefs <- coef(model)

  out <- pmap(pairs, difference_pointwise, smooth = smooth, by_var = by_var,
              smooth_var = smooth_var, data = sm_data, Xp = Xp, V = V,
              coefs = coefs, nrep = nrep)
  out <- bind_rows(out)
  crit <- qnorm((1 - ci_level) / 2, lower.tail = FALSE)

  out <- add_column(out,
                    lower = out$diff - (crit * out$se),
                    upper = out$diff + (crit * out$se),
                    .after = 6L)
  out
}

#compute difference between smooths and calculate confidence interval: complete data
diff_complete <- difference_smooths(gam_02, smooth = "s(Day)", newdata = newdat,
                     unconditional = TRUE, frequentist = FALSE,
                     n=100, partial_match = TRUE, nrep=10000)


#function to obtain values for the shading regions
pairwise_limits<-function(dataframe){
    #extract values where the lower limit of the ribbon is greater than zero
    #this is the region where the control group effect is greater

    v1<-dataframe%>%
        filter(lower_s>0)%>%
        select(Day)
    #get day  initial value
    init1=v1$Day[[1]]
    #get day final value
    final1=v1$Day[[nrow(v1)]]
    #extract values where the value of the upper limit of the ribbon is lower than zero
    #this corresponds to the region where the treatment group effect is greater
    v2<-dataframe%>%
        filter(upper_s<0)%>%
        select(Day)

    init2=v2$Day[[1]]
    final2=v2$Day[[nrow(v2)]]
    #store values
    my_list<-list(init1=init1,
                  final1=final1,
                  init2=init2,
                  final2=final2)
    return(my_list)

    

}

my_list<-pairwise_limits(diff_complete)
rib_col<-'#8D7D82' #color for ribbon for confidence interval
control_rib <- '#875F79' #color for ribbon for control region
treat_rib <- '#A7D89E' #color for ribbon treatment region

c1 <- ggplot() +
  geom_line(data = diff_complete, aes(x = Day, y = diff),size=1, alpha=0.5) +
    annotate("rect",
                xmin =my_list$init1, xmax =my_list$final1,ymin=-Inf,ymax=Inf,
                fill=control_rib,
                alpha = 0.5,
                )+
    annotate("text",
             x=1.5,
             y=-18,
             label="Control>Treatment",
           size=6,
           angle=90
           )+
    annotate("rect",
             xmin =my_list$init2, xmax =my_list$final2,ymin=-Inf,ymax=Inf,
             fill=treat_rib,
             alpha = 0.5
             ) +
  annotate("text",
             x=6,
             y=-18,
             label="Treatment>Control",
             size=6,
           angle=90
           )+
  geom_ribbon(data = diff_complete, aes(x = Day, ymin = lower_s, ymax = upper_s),
              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
  geom_hline(yintercept = 0, lty = 2, color = "red")+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    labs(y="Difference\n(Complete observations)")+
    theme_classic()+
    theme(
    axis.text=element_text(size=22))
    
    

###compute difference in smooths for missing data
diff_missing <- difference_smooths(mod_m1, smooth = "s(Day)", newdata = newdat,
                     unconditional = TRUE, frequentist = FALSE,
                     n=100, partial_match = TRUE, nrep=10000)

v2<- diff_missing %>%
        filter(upper_s<0)%>%
        select(Day)
init2=v2$Day[[1]]
final2=v2$Day[[nrow(v2)]]

c2 <- ggplot() +
  geom_line(data = diff_missing, aes(x = Day, y = diff),size=1, alpha=0.5) +
    annotate("rect",
             xmin =init2, xmax = final2,ymin=-Inf,ymax=Inf,
             fill=treat_rib,
             alpha = 0.5,
    ) +
  annotate("text",
             x=6,
             y=-18,
             label="Treatment>Control",
             size=6,
           angle=90
           )+
  geom_ribbon(data = diff_missing, aes(x = Day, ymin = lower_s, ymax = upper_s),
              alpha = 0.5, fill = rib_col, inherit.aes = FALSE) +
  geom_hline(yintercept = 0, lty = 2, color = "red")+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    labs(y="Difference\n(Missing observations)")+
    theme_classic()+
    theme(
    axis.text=element_text(size=22))
    


pair_comp<-c1+c2
```

(ref:pairwise-comp) Pairwise comparisons for smooth terms. **A**: Pairwise comparisons for the full dataset.  **B**: Pairwise comparisons for the dataset with missing observations. Significant differences exist where the 95% empirical Bayesian credible interval does not cover 0. In both cases the effect of treatment is significant after day 3. In both cases, a simultaneous CI has been computed around the smooth difference.

```{r,plot-pairwise-comp,fig.width=12, fig.height=7, out.width='100%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp)'}

pair_comp+plot_annotation(tag_levels='A')&
    theme(
     text=element_text(size=18)
     )
```


## Determination of significance in GAMs for longitudinal data{#GAM-significance}

At the core of a biomedical longitudinal study lies the question of a significant difference between the effect of two or more treatments in different groups. Whereas in rm-ANOVA a _post-hoc_ analysis is required to answer such question by calculating some _p-values_ after multiple comparisons, GAMs can use a different approach to estimate significance. In essence, the idea behind the estimation of significance in GAMs across different treatment groups in a model where separate smoothers exists per group (such as in `gam_02`) is that the difference between them can be computed, followed by the estimation of a confidence interval around this difference. 

The absence of a _p-value_ in this context might seem odd, but its interpretation can be conceptualized in the following manner: Different time trends in each group are an indication of an effect by the treatment. This is what happens for the simulated data in Figure \@ref(fig:sim-smooth-plot)A, where the chemotherapy causes $\mbox{StO}_2$ to increase over time. 

With this expectation of different trends in each group, computing the difference between the trends will identify if the observed difference is significant. A difference between groups with similar trends is unlikely to be distinguishable from zero, which would indicate that the treatment is not causing a change in the response in one of the groups (assuming the other group is a Control or Reference group).

Consider the calculation of pairwise differences for the smooths in Figure \@ref(fig:sim-smooth-plot)B and Figure \@ref(fig:sim-smooth-plot)D. Figure \@ref(fig:plot-pairwise-comp) shows the comparison between each treatment group for the full and missing datasets with a simultaneous CI computed around the difference. Here, the  "Control" group is used as the reference to which "Treatment" group is being compared. Notice that because we have included the means of each group, there is correspondence between the scale of the original data the scale of the pairwise comparisons. This can be seen as in Figure \@ref(fig:sim-smooth-plot)B at day 5 there is essentially a 50% difference between $\mbox{StO}_2$ in both groups, which corresponds to the -50 difference in \@ref(fig:plot-pairwise-comp)A. However, if there are multiple parametric terms in the model (more factors that need to be specified in the model) such inclusion of the means can become problematic. However, we believe that the model we have presented here suffices in a wide range of situations where adding the group means is a relatively easy implementation that can help better visualize the estimation from the model from a biological perspective.

In Figure \@ref(fig:plot-pairwise-comp)A the shaded regions over the confidence interval (where it does not cover 0) indicate the time interval where each group has a higher effect than the other. Notice that the shaded region between days 0 and $\approx$ 2 for the full dataset indicates that through that time, the "Control" group has higher mean $\mbox{StO}_2$, but as therapy progresses the effect is reversed and by $\approx$ 3 day  it is the "Treatment" group the one that on average, has greater $\mbox{StO}_2$. This would suggest that the effect of chemotherapy in the "Treatment" group becomes significant after day 3 for the given model. Moreover, notice that although there is no actual measurement at day 3, the model is capable of providing an estimate of when the shift in mean $\mbox{StO}_2$ occurs. 

On the data with missing observations (Figure \@ref(fig:sim-smooth-plot)D), the smooth pairwise comparison (Figure \@ref(fig:plot-pairwise-comp)B) shows that because the confidence intervals overlap for the first two days there is essentially no difference between the groups. However, because the model was still able to pick the overall trend in $\mbox{StO}_2$, the pairwise comparison is able to estimate the change on day 3 where the Treatment Group becomes significant as in the full dataset smooth pairwise comparison. 

In a sense, the pairwise smooth comparison is more informative than a _post-hoc_ _p-value_. For biomedical studies, the smooth comparison is able to provide an estimate of _when_ and by _how much_ a biological process becomes significant. This is advantageous because it can help researchers gain insight on metabolic changes and other biological processes that can be worth examining, and can help refine the experimental design of future studies in order to obtain measurements at time points where a significant change might be expected.

