---
title: '**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**'
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
author:
- Ariel Mundo^[Department of Biomedical Engineering, University of Arkansas, Fayetteville]
- Timothy J. Muldoon^[Department of Biomedical Engineering, University of Arkansas,
  Fayetteville]
- John R. Tipton^[Department of Mathematical Sciences, University of Arkansas, Fayetteville]
output:
  bookdown::word_document2:
    fig_caption: yes
    keep_md: yes
  bookdown::pdf_document2:
    keep_tex: yes
    fig_caption: yes
    extra_dependencies: ["subfig","breqn"]
  bookdown::html_document2: default
  link-citations: yes
  css: style.css
  '': default
csl: Elsevier.csl
bibliography: refs.bib
---

# Background

A longitudinal study is designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study). Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect in patients before, during and after therapy. For example, Sio et. al. analyzed the evolution of dermatitis symptoms in breast cancer patients following radiotherapy (RT), by taking weekly measurements over the course of two months [@sio2016], and Kamstra et al measured mouth opening at regular monthly intervals following RT in neck cancer patients [@kamstra2015]. Other studies using longitudinal designs in biomedical research have tumor  analyzed tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010]. 


Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as a limiting frequency[@wagenmakers2008] by assuming a null hypothesis under a statistical model that is often assumed to be an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model typically makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018]. 

Indeed, an expected linear behavior of the response through time is a requisite in rm-ANOVA [@pinheiro2006]. This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model can only accommodate linear trends.

Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric relies on how adequate the model fits the data. A _post hoc_ analysis is based on multiple repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance. The _p-value_ requires that the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019]. For example, if a model that is only able to explain linear responses is fitted to data that follows a quadratic trend, the resulting _p-values_ and parameter estimates  are invalid[@wang2019].

Additionally, the _p-value_ is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study.Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019]. Therefore, the _post hoc_ comparisons commonly used in rm-ANOVA impose a tradeoff between Type I, II and III errors.


The assumption of constant correlation (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004]. 


Due to a variety of causes, the number of observations during a study can vary between all subjects. For example, in a clinical trial patients may voluntarily withdraw, whereas attrition  due to injury or weight loss in preclinical animal studies is a possible and indeed common occurrence. It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points. In each of these missing data scenarios,the _complete observations_ assumption of  classical rm-ANOVA is violated. When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis[@gueorguieva2004]. This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups. 

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information. The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor data [@skala2010;@vishwanath2009]. Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006]. On the other hand, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013]; but more importantly, they also expect a linear relationship between the response and time [@pinheiro2006], making them also limited to analyze non-linear data.

As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropiate inference.

In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not frequently used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] . There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations. Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.

The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs. In particular, $\textsf{R}$[@r]has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g. hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017]. Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and can be also used for power calculations and study design questions.   

This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations is explained in detail. Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations. Third, using simulated data that reproduces patterns in previously reported studies [@vishwanath2009] we illustrate the type of non-linear longitudinal data that often occurs in biomedical research. The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies. Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in $\textsf{R}$, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.  

In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models by improving the standards for reproducibility  in biomedical research.


# Challenges presented by longitudinal studies

## The repeated measures ANOVA

The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research, but certain assumptions are necessary for the model to be valid. From a practical view, the assumptions can be divided in three areas: a linear relationship between covariates and response, a constant correlation between measurements, and complete observations for all subjects. Each one of these assumptions is discussed below.
  
## Linear relationship 

In a biomedical longitudinal study, two or more groups of subjects (patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs. a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points. The collected respnse is modeled with both  _fixed_ and  _random_ components. The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the average effect of the novel drug in the "treatment" group. The _random_ component can be defined as "noise" caused by some inherent variability within the study. For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that the difference in the collection time of the measurement introduces some "noise" in the signal. As their name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.


Mathematically speaking, a rm-ANOVA model with an interaction can be written as:

\begin{equation}
y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
(\#eq:linear-model)
\end{equation}

In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively. The _random effect_ $\mu_{ij}$ is assumed to be independently normally distributed with mean zero and variance $\sigma^2_\mu$, which can be expressed as $\mu_{ij} \sim N(0,\sigma^2_\mu)$. This _random effect_ accounts for variability in each subject within each group by allowing different values for each subject. Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ or _random_ effects, and are assumed to be $\sim N(0,\sigma^2)$ . 
In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs. "novel drug" or "saline" vs. "chemotherapy"). Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B). The linear models then can be expressed as


\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\beta_0 + \beta_1 \times time_{t} + \beta_2 \times +\beta_3 \times time_{t} \times +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-by-group)
\end{equation}

To further simplify the expression, substitute $\widetilde{\mu}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B. This substitution allows for a different intercept and slope for Groups A and B. The model is then written as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\widetilde{\mu} + \widetilde{\beta_1} \times time_{t} +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-lines)
\end{equation}

Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group). In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data. If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines. Because the construction of the model is essentially the same in LMEMs, the linearity assumption holds for these models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].  

## Covariance in rm-ANOVA and LMEMs

In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group. The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]). 

In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity"). Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976]. However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004]. Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001]. Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.


In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006]. Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a long process by itself. Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis. 


## Missing observations

Missing observations are an issue that arises frequently in longitudinal studies. In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004]. Dropout from patients and attrition or injury in animals are among the reasons for missing observations. Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].  In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002]. If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value. An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].

In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014]. For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.

This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data. Of notice, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study. However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data. To better convey the issues of linearity and correlation  in linear models fitted to non-linear data, simulation is used in the next section.

## Simulation experiment on linearity and covariance structures

To demonstrate the limitations of rm-ANOVA for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each. An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix). 
Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:linear-models, A)). In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:quadratic-response),A). 

The rationale for the chosen linear and quadratic functions is the expectation that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the response indicates a difference in the effect of each treatment. From the mean response per group (linear or quadratic), individual responses for each subject within each group are simulated using covariance matrices with compound symmetry (constant variance across time) or independent errors (errors generated from a a normal distribution that are not constant over time). Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (either compound symmetry or independent, which appear in panels A and E, respectively in both Figure \@ref(fig:linear-models) and Figure \@ref(fig:quadratic-response)) . The resulting simulated data, simulated errors, individual response and fitted parameters from the model are plotted.

```{r, message = FALSE,include=FALSE}
set.seed(11)
library(patchwork)
library(tidyverse)
```


```{r,include=FALSE,message=FALSE,echo=FALSE}
## Example with linear response
example <- function(n_time = 6, 
                    fun_type = "linear", 
                    error_type = "correlated") {
  
  if (!(fun_type %in% c("linear", "quadratic")))
    stop('fun_type must be either "linear", or "quadratic"')
  if (!(error_type %in% c("correlated", "independent")))
    stop('fun_type must be either "correlated", or "independent"')
  
  library(tidyverse)
  library(mvnfast)
  library(nlme)
  library(ggsci)
  
  x <- seq(1,6, length.out = n_time)
  mu <- matrix(0, length(x), 2)
  # linear response
  if (fun_type == "linear") {
    mu[, 1] <- - (0.25*x)+2  
    mu[, 2] <- 0.25*x+2
  } else {
    # nonlinear response
    
    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
  }
  # matplot(mu, type = 'l')
  
  y <- array(0, dim = c(length(x), 2, 10))
  errors <- array(0, dim = c(length(x), 2, 10))
  
  if (error_type == "independent") {
    ## independent errors
    for (i in 1:2) {
      for (j in 1:10) {
        errors[, i, j] <- rnorm(6, 0, 0.25)
        y[, i, j] <- mu[, i] + errors[, i, j]
      }
    }
  } else {
    for (i in 1:2) {     # number of treatments
      for (j in 1:10) {  # number of subjects
        # compound symmetry errors
        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
        y[, i, j] <- mu[, i] + errors[, i, j]
      }
    }
  }    
  
  
  ## subject random effects
  
  ## visualizing the difference between independent errors and compound symmetry
  ## why do we need to account for this -- overly confident inference
  
  
  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
  dimnames(mu) <- list(time = x, treatment = 1:2)
  dat <- as.data.frame.table(y, responseName = "y")
  dat_errors <- as.data.frame.table(errors, responseName = "errors")
  dat_mu <- as.data.frame.table(mu, responseName = "mu")
  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
  dat$time <- as.numeric(as.character(dat$time))
  dat <- dat %>%
    mutate(subject = factor(paste(subject, treatment, sep = "-")))
  
  
  ## repeated measures ANOVA in R
  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
  dat$preds_lm <- predict(fit_lm)

  fit_lme <- lme(y ~ treatment + time + treatment:time,
                 data = dat,
                 random = ~ 1 | subject,
                 correlation = corCompSymm(form = ~ 1 | subject)
  )
  
  
  pred_dat <- expand.grid(
    treatment = factor(1:2), 
    time = unique(dat$time)
  )
  
  dat$y_pred <- predict(fit_lme)
  
  
  return(list(
    dat = dat,
    pred_dat = pred_dat,
    fit_lm = fit_lm,
    fit_lme = fit_lme
    
  ))
}

plot_example <- function(sim_dat) {
  library(patchwork)
  ## Plot the simulated data
  p1 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
    geom_point(show.legend=FALSE) +labs(y='response')+
    geom_line(aes(x = time, y = mu, color = treatment),show.legend=FALSE) +
    theme_classic() +
    #ggtitle("Simulated data with true response function")+
    theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    scale_color_aaas()
  
  p2 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
    # facet_wrap(~ treatment) +
    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"), lty = 1,show.legend = FALSE) +labs(y='response')+
    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
    #ggtitle("Simulated data\nIndividual responses with population mean") +
    theme_classic()+
     theme(plot.title = element_text(size = 30, 
                                face = "bold"),
     text=element_text(size=30))+
    scale_color_aaas()
  
   p3 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
    geom_line(show.legend=FALSE) +labs(y='errors')+
     theme_classic()+
    # facet_wrap(~ treatment) +
    #ggtitle("Simulated errors") +
     theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    scale_color_aaas()
  
  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
    geom_point()+labs(y='response')+
    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    #ggtitle("Fitted Model")+
    theme(plot.title = element_text(size = 30, 
                                  face = "bold"),
        text=element_text(size=30))+
    scale_color_aaas()
  
  return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A'))
  # return((p1 + p3 + plot_layout(guides = "collect")) / (p2 + p4 + plot_layout(guides = "collect")))

}

txt<-18
A<-plot_example(example(fun_type = "linear", error_type = "correlated")) 


B<-plot_example(example(fun_type = "linear", error_type = "independent")) 
  
C<-plot_example(example(fun_type = "quadratic", error_type = "correlated")) 
  

D<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
  

```

```{r, linear-models, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model. A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data. B,F: Generated errors showing the difference in the behavior of correlated and independent errors.  C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline. D,H: Estimations from the rm-ANOVA model for the mean group response. Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
# linear response, correlated errors (subject effect)
A/B+plot_annotation(tag_levels = 'A')
```

In Figure \@ref(fig:linear-models)  panels C and G show a visual interpretation of "correlation" in the responses: In panel C, subjects that have a value of the random error $\varepsilon$ either above or below the mean group response are more likely to have other observations that follow the same trajectory, thereby demonstrating correlation in the response. In panel G,because the errors are independent, there is no expectation that responses are likely to follow a similar pattern.  It is clear from Figure \@ref(fig:linear-models) that the fit produced by the rm-ANOVA model is good, as the predictions for the mean response are reasonably close to the "truth" of the simulated data.

However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:quadratic-response). Here, the mean response per group was simulated using a quadratic function and errors, individual responses and the rm-ANOVA model were produced in the same manner as \@ref(fig:linear-models) . The figure shows that the response is changing in each group through the timeline, and that the mean value is the same as the initial value by the fifth time point. Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panels D and H in Figure \@ref(fig:quadratic-response).

```{r, quadratic-response, fig.width=16, fig.height=12,  echo=FALSE,message=FALSE,fig.show='hold',fig.cap="**Simulated quadratic responses from two groups with a rm-ANOVA model fitted. A,E:Simulated data with known mean response (lines) and individual responses (points) showing the dispersion of the data. B,F: Generated errors showing the difference in the behavior of correlated and independent errors.  C,G: Simulated known response per group (thick lines) with individual trajectories (thin lines), note that subjects with observations in the area above the mean response tend to stay in that region through the timeline. D,H: Estimations from the rm-ANOVA model for the mean group response. Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data.**"}
# linear response, correlated errors (subject effect)
C/D+plot_annotation(tag_levels='A')
```


In this case, Figure \@ref(fig:quadratic-response), panel A shows that both groups start with the same mean value at the first time point. Again, this is intentional in order to simulate the temporal behavior of some physiological quantity. Therefore, the idea behind \@ref(fig:quadratic-response) is to mimic a non-linear behavior from two groups that follow an inverse trajectory.  A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data (\@ref(fig:quadratic-response), panels D and H)  indicates that the model is not capturing the changes within each group in a good way.  Specifically, note that in Figure \@ref(fig:quadratic-response) the fitted mean response of the rm-ANOVA model (panels D and H) show that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model and moreover, the model is not being able to capture the initial value (which is the same in each group), instead fitting non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels A and D, and E and H). If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response, potentially limiting valuable inferences from the data. 

This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data. Although the model fitted to the simulated data was an rm_ANOVA, the main issue of an expected linear trend in the response is the same in the case of LMEMs. In the following section, we present generalized additive models (GAMs) as an alternative to analyze longitudinal non-linear data.

# GAMs as a special case of Generalized Linear Models

## GAMs and Basis Functions

Generalized linear models (GLMs) are a family of models that fit a linear response function to data that do not have normally distributed errors[@nelder1972]. In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case[@simpson2018;@wood2017;@hastie1987]. A GAM model can be written as:


\begin{equation}
  y_{ijt}=\beta_0+f(x_t\mid \beta_j)+\varepsilon_{ijt}
  (\#eq:GAM)
\end{equation}

Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ and $\varepsilon_{ijt}$ represents the residual error.

In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_. This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response. One possible function for $f(x_t\mid \beta_j)$ that allows for non-linear responses is a polynomial, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998]. In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.


The smooth functional relationship between the covariates and the response in GAMs is specified   using a parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions and this step is achieved by using _basis functions_ to represent them. A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018]. For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$. The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.  In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows relationships beyond linear for the covariates.

```{r,basis-functions}
library(mgcv)
library(mvnfast)
  library(nlme)
  library(ggsci)
n_time = 6
 x <- seq(1,6, length.out = n_time)
 mu <- matrix(0, length(x), 2)
 mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
 mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
 y <- array(0, dim = c(length(x), 2, 10))
 errors <- array(0, dim = c(length(x), 2, 10))
 for (i in 1:2) {     # number of treatments
     for (j in 1:10) {  # number of subjects
         # compound symmetry errors
         errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
         y[, i, j] <- mu[, i] + errors[, i, j]
     }
 }
  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
 dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
 dimnames(mu) <- list(time = x, treatment = 1:2)
 dat <- as.data.frame.table(y, responseName = "y")
 dat_errors <- as.data.frame.table(errors, responseName = "errors")
 dat_mu <- as.data.frame.table(mu, responseName = "mu")
 dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
 dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
 dat$time <- as.numeric(as.character(dat$time))
 dat <- dat %>%
     mutate(subject = factor(paste(subject, treatment, sep = "-")))
  
  dat<-subset(dat,treatment==1)
  dat<-dat[,c('y','time')]
gm<-gam(y~s(time,k=5),data=dat)
model_matrix<-predict(gm,type='lpmatrix')

x_new <- (seq(0, max(dat$time), length.out = 100))
y_pred <- predict(gm, newdata=dat) 

time<-c(1:6)

basis1<-model_matrix[1:6,2]
basis2<-model_matrix[1:6,3]
basis3<-model_matrix[1:6,4]
basis4<-model_matrix[1:6,5]

plot(time,basis1,type='l')
plot(time,basis2,type='l')
plot(time,basis3,type='l')
plot(time,basis4,type='l')


#plot(dat$y~dat$time)

#matplot(dat$time, model_matrix[,-1], type = "l", lty = 2, add = T)

#lines(y_pred ~ dat$time, col = "red", lwd = 2)

```


In GAMs the _basis functions_ are typically evenly spaced in _knots_ over the range of the covariates. A commonly used _basis function_ is the cubic spline, which is a smooth curve constructed from cubic polynomials joined together at the knot locations[@wood2017;@simpson2018]. Cubic splines have a long history in solving non-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain visual smoothness [@wegman1983]. Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.

The use of GAMs to analyze biomedical longitudinal data, and the impact of missing observations in the fit of the model will be examined in detail in the following section using simulation.



***



# References
