
# Challenges presented by longitudinal studies

## The repeated measures ANOVA

The _repeated measures analysis of variance_ (rm-ANOVA) is the standard statistical analysis for longitudinal data in biomedical research. This statistical methodology requires certain assumptions for the model to be valid. From a practical view, the assumptions can be divided in three areas: 1) linear relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects. Each one of these assumptions is discussed below.
  
## Linear relationship 

### The repeated measures ANOVA case

In a longitudinal biomedical study, two or more groups of subjects (e.g., human subject, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug or intervention vs. a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points. The collected response is modeled with _fixed_ components. The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug/intervention in the "treatment" group. 


Mathematically speaking, a rm-ANOVA model with an interaction can be written as:

\begin{equation}
y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
(\#eq:linear-model)
\end{equation}

In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively. Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$ (independently normally distributed with mean zero and variance $\sigma^2_\mu$).
In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs. "novel drug" or "saline" vs. "chemotherapy"). Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B). The linear models then can be expressed as


\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\beta_0 + \beta_2+\beta_1 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-by-group)
\end{equation}

To further simplify the expression, substitute $\widetilde{\beta_{0}}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B. This substitution allows for a different intercept and slope for Groups A and B. The model is then written as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-lines)
\end{equation}

Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group). In other words, the rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data. If the data show non-linear behavior, the rm-ANOVA model will approximate this behavior with non-parallel lines. 

### The Linear Mixed Model Case

A linear mixed model (LMEM) is a class of statistical model that incorporates _fixed effects_ to model the relationship between the covariates and the response, and _random effects_ to model subject variability that is not the primary focus of the study but that might be important to distinguish [@pinheiro2006;@west2014]. A LMEM with interaction between time and treatment for a longitudinal study  can be written as:


\begin{equation}
y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{ijt}\\ 
(\#eq:LMEM)
\end{equation}

When Equation \@ref(eq:linear-model) and Equation \@ref(eq:LMEM) are compared, it is easily noticeable that LMEM and rm-ANOVA have the same construction regarding the _fixed effects_ of time and treatment, but that the LMEM incorporates an additional source of variation (the term $\mu_{ij}$). This term $\mu_{ij}$ is the one that corresponds to the  _random effect_, accounting for variability in each subject within each group. The _random_ component can also be understood as used to model some "noise" in the response, but that is intended to be analyzed and disentangled from the "global noise" term $\varepsilon_{ijt}$ from Equation \@ref(eq:linear-model). 

For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some "noise" in the data. As the name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.  The _random effect_ $\mu_{ij}$ in Equation \@ref(eq:LMEM) is assumed to be $\mu_{ij} \sim N(0,\sigma^2_\mu)$. In essence,the _random effect_ in a LMEM enables to fit models with different slopes at the subject-level[@pinheiro2006]. However,the expected linear relationship of the covariates and the response in Equation \@ref(eq:linear-model) and in Equation \@ref(eq:LMEM) is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.

## Covariance in rm-ANOVA and LMEMs

In a longitudinal study there is an expected _covariance_ between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a _covariance_ between  measurements at each time point within each group. The _covariance matrix_ (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study[@wolfinger1996] (For an in-depth analysis of the covariance matrix see [@west2014;@weiss2005]). 

In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as _compound symmetry_ (also known as "sphericity" or "circularity"). Under this assumption, the between-subject variance and within-subject correlation  are constant across time [@weiss2005;@geisser1958;@huynh1976]. However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time [@maxwell2017]; and it is higher between consecutive measurements [@gueorguieva2004;@ugrinowitsch2004]. Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser)[@huynh1976;@greenhouse1959] the effectiveness of each correction is limited because it depends on the size of the sample,the number of repeated measurements[@haverkamp2017], and they are not robust if the group sizes are unbalanced [@keselman2001]. Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.


In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others [@pinheiro2006]. Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself. Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis. 


## Missing observations

Missing observations are an issue that arises frequently in longitudinal studies. In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004]. Dropout from patients and attrition or injury in animals are among the reasons for missing observations. Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].  In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002]. If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value. An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].

In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014]. For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.

This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data. In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study. However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data. To better convey the issues of linearity and correlation  in linear models fitted to non-linear data, simulation is used in the next section.

## What does an rm-ANOVA fit looks like? A visual representation using simulated data {#simulation}

To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each. An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme]. 
Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A). In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C). In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point. This is intentional in order to simulate the expected temporal evolution of some physiological quantity.

Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as treatment progresses a divergence in the trend of the response indicates an effect due to treatment. In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group. From the mean response per group (linear or quadratic), the variability or "error" of individual responses  within each group is simulated using a covariance matrix with compound symmetry (constant variance across time). Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D). A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal data is in Figure \@ref(fig:linear-cases) and Figure \@ref(fig:quadratic-cases) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.



```{r,include=FALSE,message=FALSE,echo=FALSE}
## Example with linear response. This function generates either linear or quadratic mean responses with correlated or uncorrelated errors and fits a linear model to the data.
example <- function(n_time = 6, 
                    fun_type = "linear", 
                    error_type = "correlated") {
  
  if (!(fun_type %in% c("linear", "quadratic")))
    stop('fun_type must be either "linear", or "quadratic"')
  if (!(error_type %in% c("correlated", "independent")))
    stop('fun_type must be either "correlated", or "independent"')
  
  
  x <- seq(1,6, length.out = n_time)
  mu <- matrix(0, length(x), 2)
  # linear response
  if (fun_type == "linear") {
    mu[, 1] <- - (0.25*x)+2  
    mu[, 2] <- 0.25*x+2
  } else {
    # nonlinear response
    
    mu[, 1] <-  -(0.25 * x^2) +1.5*x-1.25
    mu[, 2] <- (0.25 * x^2) -1.5*x+1.25
  }
  # matplot(mu, type = 'l')
  
  y <- array(0, dim = c(length(x), 2, 10))
  errors <- array(0, dim = c(length(x), 2, 10))
  
  if (error_type == "independent") {
    ## independent errors
    for (i in 1:2) {
      for (j in 1:10) {
        errors[, i, j] <- rnorm(6, 0, 0.25)
        y[, i, j] <- mu[, i] + errors[, i, j]
      }
    }
  } else {
    for (i in 1:2) {     # number of treatments
      for (j in 1:10) {  # number of subjects
        # compound symmetry errors
        errors[, i, j] <- rmvn(1, rep(0, length(x)), 0.1 * diag(6) + 0.25 * matrix(1, 6, 6))
        y[, i, j] <- mu[, i] + errors[, i, j]
      }
    }
  }    
  
  
  ## subject random effects
  
  ## visualizing the difference between independent errors and compound symmetry
  ## why do we need to account for this -- overly confident inference
  
  
  dimnames(y) <- list(time = x, treatment = 1:2, subject = 1:10)
  dimnames(errors) <- list(time = x, treatment = 1:2, subject = 1:10)
  dimnames(mu) <- list(time = x, treatment = 1:2)
  dat <- as.data.frame.table(y, responseName = "y")
  dat_errors <- as.data.frame.table(errors, responseName = "errors")
  dat_mu <- as.data.frame.table(mu, responseName = "mu")
  dat <- left_join(dat, dat_errors, by = c("time", "treatment", "subject"))
  dat <- left_join(dat, dat_mu, by = c("time", "treatment"))
  dat$time <- as.numeric(as.character(dat$time))
  dat <- dat %>%
    mutate(subject = factor(paste(subject, treatment, sep = "-")))
  
  
  ## repeated measures ANOVA in R
  fit_lm <- lm(y ~ time + treatment + time * treatment, data = dat)
  dat$preds_lm <- predict(fit_lm)

  fit_lme <- lme(y ~ treatment + time + treatment:time,
                 data = dat,
                 random = ~ 1 | subject,
                 correlation = corCompSymm(form = ~ 1 | subject)
  )
  
  
  pred_dat <- expand.grid(
    treatment = factor(1:2), 
    time = unique(dat$time)
  )
  
  dat$y_pred <- predict(fit_lme)
  
  
  return(list(
    dat = dat,
    pred_dat = pred_dat,
    fit_lm = fit_lm,
    fit_lme = fit_lme
    
  ))
}

plot_example <- function(sim_dat) {
  library(patchwork)
  ## Plot the simulated data
  p1 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = y, group = treatment, color = treatment)) +
    geom_point(show.legend=FALSE,size=1.5,alpha=0.5) +
    labs(y='response')+
    geom_line(aes(x = time, y = mu, color = treatment,size=1),show.legend=FALSE) +
    theme_classic() +
    #ggtitle("Simulated data with true response function")+
    theme(plot.title = element_text(size = 18, 
                                  face = "bold"),
        text=element_text(size=18))+
    scale_color_aaas()
  
  p2 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = y, group = subject, color = treatment)) +
    geom_line(aes(size = "Subjects"),show.legend = FALSE) +
    # facet_wrap(~ treatment) +
    geom_line(aes(x = time, y = mu, color = treatment, size = "Simulated Truth"),
              lty = 1,show.legend = FALSE) +labs(y='response')+
    scale_size_manual(name = "Type", values=c("Subjects" = 0.5, "Simulated Truth" = 3)) +
    #ggtitle("Simulated data\nIndividual responses with population mean") +
    theme_classic()+
     theme(plot.title = element_text(size = 18, 
                                face = "bold"),
     text=element_text(size=18))+
    scale_color_aaas()
  
   p3 <- sim_dat$dat %>%
    ggplot(aes(x = time, y = errors, group = subject, color = treatment)) +
    geom_line(show.legend=FALSE) +labs(y='errors')+
     theme_classic()+
     theme(plot.title = element_text(size = 18, 
                                  face = "bold"),
        text=element_text(size=18))+
    scale_color_aaas()
  
  p4 <- ggplot(sim_dat$dat, aes(x = time, y = y, color = treatment)) +
    geom_point(size=1.5,alpha=0.5)+labs(y='response')+
    geom_line(aes(y = predict(sim_dat$fit_lme), group = subject, size = "Subjects")) +
    geom_line(data = sim_dat$pred_dat, aes(y = predict(sim_dat$fit_lme, level = 0, newdata = sim_dat$pred_dat), size = "Population")) +
    scale_size_manual(name = "Predictions", values=c("Subjects" = 0.5, "Population" = 3)) +
    theme_classic() +
    #ggtitle("Fitted Model")+
    theme(plot.title = element_text(size = 18, 
                                  face = "bold"),
        text=element_text(size=18))+
    scale_color_aaas()
  
  #return((p1+p3+p2+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A')) 
  
  return((p1+p4)+plot_layout(nrow=1)+plot_annotation(tag_levels = 'A')) 
    
}

txt<-18
A<-plot_example(example(fun_type = "linear", error_type = "correlated")) 

B<-plot_example(example(fun_type = "linear", error_type = "independent")) 
  
C<-plot_example(example(fun_type = "quadratic", error_type = "correlated")) 
  
D<-plot_example(example(fun_type = "quadratic", error_type = "independent")) 
  

```

(ref:l-q-response-caption) Simulated linear responses from two groups with correlated (top row) or independent (bottom row) errors using a rm-ANOVA model. A, C:Simulated data with known mean response (linear or quadratic, thin lines) and individual responses (points) showing the dispersion of the data. B,D: Estimates from the rm-ANOVA model for the mean group response (linear of quadratic). Thick lines are the predicted mean response per group, thin lines are the random effects for each subject and points represent the original raw data. The rm-ANOVA model not only fails to pick the trend of the quadratic data but it also incorrectly estimates the initial conditions.

```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='ref(l-q-response-caption)' }
# linear response, correlated errors (subject effect)
par(mar = c(0.5, 0.5, 0.5, 0.5))
A/C+plot_annotation(tag_levels = 'A')
```

The simulation shows that the fit produced by the rm-ANOVA model is good for linear data, as the predictions for the mean response are reasonably close to the "truth" of the simulated data  (Figure \@ref(fig:l-q-response),B). When the linearity and compound symmetry assumptions are met, the model approximates well the individual trends and the mean trends by group.

However, consider the case when the data follows a non-linear trend, such as the simulated data in Figure \@ref(fig:l-q-response), C. Here, the mean response per group was simulated using a quadratic function but errors, individual responses and the rm-ANOVA model were produced in the same manner as in (Figure \@ref(fig:l-q-response) A and B) . The mean response in the simulated data with quadratic behavior is changing in each group through the timeline, and the mean value is the same as the initial value by the fifth time point for each group. Fitting an rm-ANOVA model \@ref(eq:linear-model) to this data  produces the fit that appears in panel D in Figure \@ref(fig:l-q-response).

A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group in a good way.  Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model. Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D). If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response. Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.  

This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data. Although the model fitted to the simulated data was an rm-ANOVA model, the main issue of an expected linear trend in the response is the same in the case of a LMEM. In the following section, we present generalized additive models (GAMs) as a data-driven  alternative method to analyze longitudinal non-linear data.
